Declarative Pearl: Lightweight Probabilistic Programming in Curry

### 2015-08-31 Paper/paper.tex
Add ideas for each chapter

\section{Introduction}
There are many probabilistic languages around these days; a progress
that started longer than a decade ago, but is currently especially
active. %
These languages have their origin either in functional, logic or
imperative programming, i.e., you can choose a language based on your
favourite paradigm. %
Besides newly developed languages, a few libraries for probabilistic
programming have seen the light of day as well. %
These languages (stand-alone or DSL) all come with a handful of
useful tools (functions and stuff) to work with probabilistic values
and distributions as well as inference algorithms to solve
probabilistic models that use these values and distributions. %
That said, there is a large subset that all these languages have in
common, dispite their different approaches on semantic, syntax as well
as goals and range of usage, respectively. %
This subset is rather of practical nature and not of big intereset
from a PL researcher perspective: every good library or programming
language needs a handful of primitives and predefined functions, but
they are not the heart of this piece of software. %

In case of probabilistic programming languages and libraries, we need
to look at the implementation of the actual semantic behaviour -- what
can we express with probabilistic languages and how can we interpret
such an expression? %
What is the intuition behind probability in the context of programming
and what is its meaning? %

The main goal of probabilistic programming is to open the door to the
evaluation of probabilistic models for a wide range of users: data
scientists, domain experts and other developers. %
Probabilistic programming languages want to minimise the burden for
non-experts (in the area of machine learning and co) to design their
own model and run inference algorithms on their data. %
These languages already provide a rich set of primitives and most
importantly inference algorithms that are built upon the results of
experts. %
Thus, a predefined and well-elaborated toolset simplifies the access
to machine learning, because technical details can be ignored. %

@@

Experts often state that you do not need to be a rocket scientist in
order to make sense of research results; however, in the area of
machine learning, beginners often face difficulties to use the tools
and current knowledge for their own project. %
That is, beginners need to inverse many hours to digest technical
details before they can actually make use of top-of-the-art models and
algorithms. %

In the field of programming languages, the community often tries to
generalise and simplify commen concepts and repetative structures. %
Fortunately, the rising collaboration between ML and PL communities
has produced a number of programming languages and libraries for
probabilistic programming that aim at minimising the obstacle to use
practicle results of ML, thus, empower more users to apply ML
techniques to their problems. %

@@

After defining the model, we want to ask queries about certian
propoerties and probabilities, respectively. %
The essence of probabilistic programming is to harmonise the problem
and its analysis: we want to define both, the problem and the queries,
in same language. %

@@

@@

\item Working within a host language enables us to focus on the
  problem itself and frees us from reimplementing standard primitives
  (IO, data structures), essential design decisions (syntax,
  parser). %
  Based on which criteria do you choose the host language? %
  As for probabilistic languages that started growing over time,
  libraries that enable probablistic programming exist in the whole
  variety of paradigms: declarative as well as imperative. %
  For a start, you obviously choose the paradigm that has the clearest
  structure and enables straight-forward programming. %
  We want to address a wide range of developers and even
  non-PL-experts, thus, we choose a paradigm than can be learnt
  easily. %
  Thus, obvioulsy, you choose a declarative programming language. %
  By choosing Curry, we argue that it is sufficient to reuse built-in
  features in order to cover the essence of probabilistic programs,
  especially its semantic. %

@@

\section{Conclusion and Future Work}

>>>>

\section{Introduction}
\begin{itemize}
\item more and more PPL
\item all paradigms are represented
\item (E)DSLs as well
\item all languages have common ground (similar primitives and
  predefined helper functions)
\item implementation is the heart of these ``software products'' --
  what can be expressed?
\item main goal of PPLs: tool for a wide range of researchers and not
  only ML experts; minimise burden for beginners; no former background
  on inference algorithms needed, define only model and ask queries;
\end{itemize}

% There are many probabilistic languages around these days; a progress
% that started longer than a decade ago, but is currently especially
% active. %
% These languages have their origin either in functional, logic or
% imperative programming, i.e., you can choose a language based on your
% favourite paradigm. %
% Besides newly developed languages, a few libraries for probabilistic
% programming have seen the light of day as well. %
% These languages (stand-alone or DSL) all come with a handful of
% useful tools (functions and stuff) to work with probabilistic values
% and distributions as well as inference algorithms to solve
% probabilistic models that use these values and distributions. %
% That said, there is a large subset that all these languages have in
% common, dispite their different approaches on semantic, syntax as well
% as goals and range of usage, respectively. %
% This subset is rather of practical nature and not of big intereset
% from a PL researcher perspective: every good library or programming
% language needs a handful of primitives and predefined functions, but
% they are not the heart of this piece of software. %

% In case of probabilistic programming languages and libraries, we need
% to look at the implementation of the actual semantic behaviour -- what
% can we express with probabilistic languages and how can we interpret
% such an expression? %
% What is the intuition behind probability in the context of programming
% and what is its meaning? %

% The main goal of probabilistic programming is to open the door to the
% evaluation of probabilistic models for a wide range of users: data
% scientists, domain experts and other developers. %
% Probabilistic programming languages want to minimise the burden for
% non-experts (in the area of machine learning and co) to design their
% own model and run inference algorithms on their data. %
% These languages already provide a rich set of primitives and most
% importantly inference algorithms that are built upon the results of
% experts. %
% Thus, a predefined and well-elaborated toolset simplifies the access
% to machine learning, because technical details can be ignored. %

@@

\begin{itemize}
\item abstraction from actual ML problems: define only model and
  questions
\item give simple example for Bayesian Networks (Wet Grass, Alarm,
  etc) and explain functions as superficial as possible
\item point out resemblence between CPT and actual code for model
\end{itemize}
% Experts often state that you do not need to be a rocket scientist in
% order to make sense of research results; however, in the area of
% machine learning, beginners often face difficulties to use the tools
% and current knowledge for their own project. %
% That is, beginners need to invest many hours to digest technical
% details before they can actually make use of top-of-the-art models and
% algorithms. %

% In the field of programming languages, the community often tries to
% generalise and simplify commen concepts and repetative structures. %
% Fortunately, the rising collaboration between ML and PL communities
% has produced a number of programming languages and libraries for
% probabilistic programming that aim at minimising the obstacle to use
% practicle results of ML, thus, empower more users to apply ML
% techniques to their problems. %

@@

After defining the model, we want to ask queries about certian
properties and probabilities, respectively. %
The essence of probabilistic programming is to harmonise a problem and
its analysis: we want to define both, the problem and the queries, in
same language. %

@@

\begin{itemize}
\item PPL vs DSL\\
  Working within a host language enables us to focus on the
  problem itself and frees us from reimplementing standard primitives
  (IO, data structures), essential design decisions (syntax,
  parser). %
  % Based on which criteria do you choose the host language? %
  % As for probabilistic languages that started growing over time,
  % libraries that enable probablistic programming exist in the whole
  % variety of paradigms: declarative as well as imperative. %
  % For a start, you obviously choose the paradigm that has the clearest
  % structure and enables straight-forward programming. %
  % We want to address a wide range of developers and even
  % non-PL-experts, thus, we choose a paradigm than can be learnt
  % easily. %
  % Thus, obvioulsy, you choose a declarative programming language. %
  % By choosing Curry, we argue that it is sufficient to reuse built-in
  % features in order to cover the essence of probabilistic programs,
  % especially its semantic. %
\item describe underlying implementation: core data types and
  functions
\item query simple question for example from introduction
\item connect to Martin Erwig's PFP approach
\item Nondeterminism and \code{SearchTree}
\end{itemize}

@@

@@

\begin{itemize}
\item Martin Erwig (FPL)
\item Oleg Kiselyov (HANSEI)
\item ProbLog (on top of Prolog), AILog (debugger), Church (functional language)
\end{itemize}

\section{Conclusion and Future Work}
\begin{itemize}
\item SetFunctions (memory allocation; usage as ``first-class
  citizen'', i.e., check semantics)
\item ``outside'' nondeterminism
\item sampling technique for nondeterministic functions in Curry
\item reevaluate Currys semantic in constrast to Sato's distributuion semantic
\item reactivate debugger for KiCS2/PAKCS
\end{itemize}
